A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-01-31 sections/Metatheory.tex
soundness and completeness

@@

\begin{theorem}[Soundness]
If \emph{wf }$\Gm$ and $\Gm \redto \nil$, then $\exists \Om$ s.t. $\Gm\sto\Om$ and $\Om\redto\nil$.
\end{theorem}

@@

@@

@@

@@

\begin{figure}
\framebox{$\Gm\jExt\Gm'$}
\begin{gather*}
\inferrule*[right=$\mathtt{\jExt\nil}$]
    {~}{\Gm\jExt\Gm}
\qquad
\inferrule*[right=$\mathtt{\jExt solve}$]
    {|A|_\forall = 0 \quad \Gm_L,[A/\al]\Gm_R \jExt \Gm'}{\Gm_L,\al,\Gm_R \jExt \Gm'}
\qquad
\inferrule*[right=$\mathtt{\jExt \al}$]
    {\Gm_L,\al,\Gm_R \jExt \Gm'}{\Gm_L,\Gm_R \jExt \Gm'}
\end{gather*}
\caption{Worklist Extension}\label{fig:worklist_ext}
\end{figure}

>>>>

The transfer $\Gm\sto\Om$ is not a function.
On the one hand, there are infinite possibilities to instantiate an existential variable,
thus a $\Gm$ could be transferred to different $\Om$'s.
On the other hand, any valid mono-type in $\Om$ can be represented by an existential variable in $\Gm$,
thus different $\Gm$'s could be transferred to a single $\Om$.

@@

Our algorithm is sound with respect to the declarative system.
For any worklist $\Gm$ that reduces successfully,
we can find a valid transfer $\Gm\sto\Om$ that solves all the existential variables,
such that $\Om$ also reduces successfully.
\begin{theorem}[Soundness]
If \emph{wf }$\Gm$ and $\Gm \redto \nil$, then $\exists \Om$ s.t. $\Gm\sto\Om$ and $\Om\redto\nil$.
\end{theorem}

The proof proceeds by induction on derivation of $\Gm\redto\nil$.
Proper applications of lemmas \ref{lem:insert} and \ref{lem:extract} finish off the theorem.

@@

Completeness of the algorithm means that any declarative system has an algorithmic counterpart.

@@

We prove completeness by induction on the derivation of $\Om\redto\nil$.
Since the declarative worklist is reduced judgment-by-judgment,
the induction always analyse the first judgment until it is satisfied.
As the algorithmic system introduces existential variables,
a declarative rule may correspond to multiple algorithmic rules,
and thus we analyse each of the possible cases.

For subtyping, algorithmic Rules 10 and 11 worth special treatment.
When the induction reaches the $\mathtt{{\le}{\to}}$ case,
the first judgment is of shape $A_1 \to A_2 \le B_1 \to B_2$.
One of the corresponding algorithmic judgment is $\al \le A \to B$.
However, the case analysis does not imply that $\al$ is fresh in $A$ and $B$,
therefore Rule~10 cannot be applied and the proof get stuck.
The following lemma helps us out in those cases:
the success in declarative subtyping indicates the freshness of $\al$ in $A$ and $B$
in its corresponding algorithmic judgment.
In other words, the declarative system do not accept infinite types,
thus, as an example, we cannot find a mono-type $\tau$ such that $\tau\le 1\to \tau$,
which could be transferred by $\al\le 1\to\al$.

@@

A symmetric lemma holds for $A\to B \le \al$.

@@

\begin{figure}
\framebox{$\Gm\jExt\Gm'$}
\begin{gather*}
\inferrule*[right=$\mathtt{\jExt id}$]
    {~}{\Gm\jExt\Gm}
\qquad
\inferrule*[right=$\mathtt{\jExt solve}$]
    {|A|_\forall = 0 \quad \Gm_L,[A/\al]\Gm_R \jExt \Gm'}{\Gm_L,\al,\Gm_R \jExt \Gm'}
\qquad
\inferrule*[right=$\mathtt{\jExt \al}$]
    {\Gm_L,\al,\Gm_R \jExt \Gm'}{\Gm_L,\Gm_R \jExt \Gm'}
\end{gather*}
\caption{Worklist Extension}\label{fig:worklist_ext}
\end{figure}
